Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2017-12-12 manuscript/time-slice.tex
extinction figure (with ggplot palette)
New appendix

\title{Time for a rethink: disparity-through-time}
%TG: Great title first part! How about changing the other part to 
%Time for a rethink: different ways of slicing through the stratigraphic cake leads to different disparity through time results.

@@

% However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular. %TG: small caveat on that after litterature search for the disparity review paper: disparity-through-time analysis have not changed much since Foote 1992, Foote and Gould 1992 and Briggs et al. 1992 in the classic palaeobiology appro,ach...
%TG: However to not piss of people, there were several attempts to "rationalise" (Wills 2001), test (Ciampaglio 2001) and expand (Pavlinov 2011 and Gerber et al 2011) namely with univariate approaches and "dtt" analysis (Harmon et al 2003).
%TG: One of the critisisms from one of the rude reviews was this idea that age of popular method has nothing to do with its performance, by nuancing this sentence we can basically say: yeah, some people have made great advances and proposed great ideas for CORRECTING ERRORS in the methods from the 90s but somehow everybody have ignored them (in palaeobiology).
% NC: OK so lets save ourselves some pain and not mention this at all and save it all for the bitching in the review paper!!!
% TG: agree!

@@

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{outlinesvg.pdf}
    \caption[caption title]
    {Outline of the disparity-through-time pipeline applied to each data set: 1) we used ancestral characters estimation to infer nodal character states, 2) we measured the pairwise Gower distance between the tips and nodal characters, 3) we ordinated the distance matrix using a PCoA, 4) we split that matrix using stratigraphy bins, equal bins and the time slice method, 5) we measured disparity through time for each of these methods.}
    \label{overview}
  \end{figure}

@@

%TG: Not sure if it's the "egeinvector" score or the "eigen" score. I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors). I call it ordination score for now.

@@

    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tips and the slice.
    % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
    %TG: I've clarified as above. Tell me if it's clearer. In my words: the distance between the slice and the node, like you illustrate in your presentation (nicely done by the way!)
          \begin{equation}
              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
          \end{equation}
          \begin{equation}
              p(\text{descendant}) = 1 - p(\text{ancestor})
          \end{equation}
    \noindent Where $d(x,y)$ is the distance measured as branch length.
    \end{enumerate}

@@

%TG: Justification on the fact that these models are not ACEs.
It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g. distance between species, variance per axis, etc.) to remain constant.

@@

\paragraph{Disparity peaks and troughs.}
We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each sub-sample. 
Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
Significant shifts in disparity peaks or troughs suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.

@@

Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). 
Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
Results for all six methods can be found in the appendix. % add details of appendix

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods. (Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}). 
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not see in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

@@

\paragraph{Disparity peaks and troughs.}
% Plots showing the differences

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
    \label{figure:peak1}
  \end{figure}

\paragraph{Effects of mass extinction events.}
% Big heat map table thing

>>>>

\title{Time for a rethink: time sub-sampling methods in disparity-through-time analyses}
%TG: Great title first part! How about changing the other part to 
%Time for a rethink: different ways of slicing through the stratigraphic cake leads to different disparity through time results.

@@

@@

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/outlinesvg.pdf}
    \caption[Outline of the disparity-through-time pipeline]
    {1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
    \label{overview}
  \end{figure}

@@

@@

    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tip at the ends of the branch and the slice.

          \begin{equation}
              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
          \end{equation}
          \begin{equation}
              p(\text{descendant}) = 1 - p(\text{ancestor})
          \end{equation}
    \noindent Where $d(x,y)$ is the distance measured as branch length.
    \end{enumerate}

@@

It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g. distance between species, variance per axis, etc.) to remain constant.

@@

\paragraph{Disparity peaks.}
We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will be influenced by the number of taxa (and/or nodes) included in each sub-sample. 
Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
We calculated the maximum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
Significant shifts in disparity peaks suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.

@@

Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). % add details of appendix
Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
Results for all six methods can be found in the appendix. % add details of appendix

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods. (Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}). 
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

@@

\paragraph{Disparity peaks}
In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}, Appendix figure). % appendix details
This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}, Appendix figure).
Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}, Appendix figure).

% Plots showing the differences
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
    \label{figure:peak1}
  \end{figure}

\paragraph{Effects of mass extinction events.}
Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}, appendix). 
However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}, appendix). % appendix!!!

% Big heat map table thing
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Extinction four example datasets.]
    {XXX}
    \label{figure:extinction1}
  \end{figure}
